---
title: "Analyzing Music Data"
author: "Oluchi Nnadi"
format: html
execute:
  echo: false
---

```{r}
#| message: false
library(tidyverse)
library(ggplot2)
library(dplyr)
library(stringr)
```

```{r}
#| cache: true
# Load the billboard dataset
data("billboard")

# 1. Prepare data
tidy_billboard <- billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  mutate(week = as.numeric(str_remove(week, "wk")))

# Get Top 10 songs
top_10_songs <- tidy_billboard |>
  group_by(artist, track) |>
  filter(min(rank) <= 10) |>
  ungroup() |>
  mutate(
    # Create a highlight label
    highlight_label = case_when(
      track == "Music" & artist == "Madonna" ~ "Fastest to #1",
      track == "Amazed" & artist == "Lonestar" ~ "Slowest to #1",
      track == "Higher" & artist == "Creed" ~ "Longest Top-10 run",
      TRUE ~ "Other Top-10"
    )
  )

# 2. Plot
ggplot(top_10_songs, aes(x = week, y = rank, group = interaction(artist, track))) +
  # Draw all top-10 songs in gray first
  geom_line(data = filter(top_10_songs, highlight_label == "Other Top-10"), 
            color = "gray80", alpha = 0.5) +
  # Draw the notable songs in color on top
  geom_line(data = filter(top_10_songs, highlight_label != "Other Top-10"), 
            aes(color = highlight_label), linewidth = 1) +
  scale_y_reverse() +
  scale_color_manual(values = c("Fastest to #1" = "#E41A1C", 
                                "Slowest to #1" = "#377EB8", 
                                "Longest Top-10 run" = "#4DAF4A")) +
  labs(
    title = "Trajectories of Billboard Top-10 Songs",
    subtitle = "Highlighting notable chart performers",
    x = "Week",
    y = "Rank (Lower is better)",
    color = "Notable Songs"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
# Read the music CSV file
music <- read.csv("data/music.csv", stringsAsFactors = FALSE)

# Select the requested numeric columns
cols <- c("artist.familiarity", "artist.hotttnesss", "song.year", "song.tempo")
selected <- music[cols]

# Convert all values to numeric and compute summary statistics
to_numeric <- function(x) {
  as.numeric(as.character(x))
}

summary_stats <- t(sapply(selected, function(x) {
  x_num <- to_numeric(x)
  x_num <- x_num[!is.na(x_num)]
  c(
    min = min(x_num),
    q1 = quantile(x_num, 0.25),
    median = median(x_num),
    q3 = quantile(x_num, 0.75),
    max = max(x_num)
  )
}))

# Print results in a readable format
print(summary_stats)
```

Artist.familiarity refers to how familiar the artist is to their listeners, with higher values indicating greater familiarity. Artist.hotttnesss measures the popularity of the artist (when downloaded), with higher values indicating greater popularity.

```{r}
# Read the music CSV file
music <- read.csv("data/music.csv", stringsAsFactors = FALSE)

# Convert song.year to numeric and filter out 0 values
music$song.year <- as.numeric(as.character(music$song.year))
music_filtered <- subset(music, song.year != 0 & !is.na(song.year))

# Count songs used in the distribution
song_count <- nrow(music_filtered)

# Create the distribution plot
library(ggplot2)
plot_year <- ggplot(music_filtered, aes(x = song.year)) +
  geom_histogram(binwidth = 5, fill = "#2c7fb8", 
  color = "white", alpha = 0.85) +
  labs(
    title = "Distribution of Song Year",
    subtitle = paste0("Number of songs used: ", song_count),
    x = "Song Year",
    y = "Count"
  ) +
  theme_minimal()

# Display the plot when interactive, otherwise save it to a PNG file
print(plot_year)
ggsave("song_year_distribution.png", plot_year, width = 10, height = 6)
  message("Plot saved to song_year_distribution.png")
```

```{r}
# Read the music CSV file
music <- read.csv("data/music.csv", stringsAsFactors = FALSE)

# Define the target columns
cols <- c(
  "artist.location",
  "release.name",
  "song.title",
  "song.year",
  "artist.familiarity",
  "artist.hotttnesss"
)

# Helper for identifying placeholders
is_placeholder <- function(x, numeric_zero = FALSE) {
  if (numeric_zero) {
    x_num <- suppressWarnings(as.numeric(as.character(x)))
    return(is.na(x_num) | x_num == 0)
  }
  x_char <- trimws(as.character(x))
  return(is.na(x_char) | x_char == "")
}

# Compute placeholder counts for each column
placeholder_counts <- sapply(cols, function(col_name) {
  col <- music[[col_name]]
  numeric_cols <- c("song.year", "artist.familiarity", "artist.hotttnesss")
  sum(is_placeholder(col, numeric_zero = col_name %in% numeric_cols))
})

# Present the results
placeholder_summary <- data.frame(
  column = cols,
  placeholder_count = placeholder_counts,
  stringsAsFactors = FALSE
)

print(placeholder_summary)
```

```{r}
# Read the music CSV file
music <- read.csv("data/music.csv", stringsAsFactors = FALSE)

# Convert coordinate columns to numeric
music$artist.latitude <- as.numeric(as.character(music$artist.latitude))
music$artist.longitude <- as.numeric(as.character(music$artist.longitude))

# Determine whether each artist row has placeholder coordinates (both latitude and longitude equal 0)
music$placeholder_coords <- with(music, artist.latitude == 0 & artist.longitude == 0)

# Count usable and placeholder coordinates
counts <- table(placeholder = music$placeholder_coords)

usable_count <- counts["FALSE"]
placeholder_count <- counts["TRUE"]

# Handle cases where one of the categories may be absent
if (is.na(usable_count)) usable_count <- 0
if (is.na(placeholder_count)) placeholder_count <- 0

# Output the results
result <- data.frame(
  category = c("usable_coordinates", "placeholder_coordinates"),
  count = c(usable_count, placeholder_count),
  stringsAsFactors = FALSE
)

print(result)
```

Many artist's location is unknown, which is why there are many placeholder coordinates. The placeholder coordinates are represented by (0, 0), which is not a valid location for any artist, so they've not been placed on the map.

```{r}
# Load required packages
# Load required packages
library(ggplot2)
library(maps)

# Read the music dataset
music <- read.csv("data/music.csv", stringsAsFactors = FALSE)

# Convert coordinate and familiarity columns to numeric
music$artist.latitude <- as.numeric(as.character(music$artist.latitude))
music$artist.longitude <- as.numeric(as.character(music$artist.longitude))
music$artist.familiarity <- as.numeric(as.character(music$artist.familiarity))

# Filter to usable coordinates and familiarity values greater than 0
usable_artists <- subset(music,
                         !(artist.latitude == 0 & artist.longitude == 0) &
                         !is.na(artist.latitude) & !is.na(artist.longitude) &
                         !is.na(artist.familiarity) & artist.familiarity > 0)

# Load world map data from maps package
world_map <- map_data("world")

# Build the plot
gg <- ggplot() +
  geom_polygon(
    data = world_map,
    aes(x = long, y = lat, group = group),
    fill = "gray95",
    color = "gray70",
    linewidth = 0.2
  ) +
  geom_point(
    data = usable_artists,
    aes(x = artist.longitude, y = artist.latitude, color = artist.familiarity),
    alpha = 0.75,
    size = 1.8
  ) +
  scale_color_viridis_c(option = "plasma", name = "Familiarity") +
  coord_fixed(1.3, xlim = c(-180, 180), ylim = c(-60, 85)) +
  labs(
    title = "Artist Locations Colored by Familiarity",
    subtitle = "Shows artists with usable coordinates and familiarity above 0; coverage is limited to geocoded artists.",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#e6f0f6"),
    panel.grid = element_line(color = "gray90"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )

# Save plot to file
print(gg)
```

```{r}
# Load required packages
library(ggplot2)
library(maps)

# Read the music dataset
music <- read.csv("data/music.csv", stringsAsFactors = FALSE)

# Convert coordinate and hotness columns to numeric
music$artist.latitude <- as.numeric(as.character(music$artist.latitude))
music$artist.longitude <- as.numeric(as.character(music$artist.longitude))
music$artist.hotttnesss <- as.numeric(as.character(music$artist.hotttnesss))

# Filter to usable coordinates and hotness values greater than 0
usable_artists <- subset(music,
                         !(artist.latitude == 0 & artist.longitude == 0) &
                         !is.na(artist.latitude) & !is.na(artist.longitude) &
                         !is.na(artist.hotttnesss) & artist.hotttnesss > 0)

# Load world map data from maps package
world_map <- map_data("world")

# Build the plot
gg <- ggplot() +
  geom_polygon(
    data = world_map,
    aes(x = long, y = lat, group = group),
    fill = "gray95",
    color = "gray70",
    linewidth = 0.2
  ) +
  geom_point(
    data = usable_artists,
    aes(x = artist.longitude, y = artist.latitude, color = artist.hotttnesss),
    alpha = 0.75,
    size = 1.8
  ) +
  scale_color_viridis_c(option = "plasma", name = "Hotttnesss") +
  coord_fixed(1.3, xlim = c(-180, 180), ylim = c(-60, 85)) +
  labs(
    title = "Artist Locations Colored by Hotttnesss",
    subtitle = "Shows artists with usable coordinates and hotttnesss above 0; coverage is limited to geocoded artists.",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#e6f0f6"),
    panel.grid = element_line(color = "gray90"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )

# Save plot to file
print(gg)
```

```{r}
# Load required packages
library(ggplot2)
library(maps)

# Read the music dataset
music <- read.csv("data/music.csv", stringsAsFactors = FALSE)

# Convert coordinate columns to numeric
music$artist.latitude <- as.numeric(as.character(music$artist.latitude))
music$artist.longitude <- as.numeric(as.character(music$artist.longitude))

# Normalize artist terms and assign genres by priority
music$artist.terms <- tolower(trimws(as.character(music$artist.terms)))
assign_genre <- function(terms) {
  if (is.na(terms) || terms == "") {
    return(NA_character_)
  }
  if (grepl("\\brock\\b", terms, ignore.case = TRUE)) {
    return("rock")
  }
  if (grepl("\\bpop\\b", terms, ignore.case = TRUE)) {
    return("pop")
  }
  if (grepl("\\bblues\\b", terms, ignore.case = TRUE)) {
    return("blues")
  }
  if (grepl("\\bjazz\\b", terms, ignore.case = TRUE)) {
    return("jazz")
  }
  if (grepl("\\bmetal\\b", terms, ignore.case = TRUE)) {
    return("metal")
  }
  return(NA_character_)
}

music$genre <- vapply(music$artist.terms, assign_genre, character(1), USE.NAMES = FALSE)

# Filter to artists with usable coordinates and a matching genre
genre_artists <- subset(music,
                        !(artist.latitude == 0 & artist.longitude == 0) &
                        !is.na(artist.latitude) & !is.na(artist.longitude) &
                        !is.na(genre) & genre != "")

# Load world map data
world_map <- map_data("world")

# Build the plot
gg <- ggplot() +
  geom_polygon(
    data = world_map,
    aes(x = long, y = lat, group = group),
    fill = "gray95",
    color = "gray70",
    size = 0.2
  ) +
  geom_point(
    data = genre_artists,
    aes(x = artist.longitude, y = artist.latitude, color = genre),
    alpha = 0.75,
    size = 1.8
  ) +
  scale_color_manual(
    values = c(
      rock = "#d73027",
      pop = "#fc8d59",
      blues = "#91bfdb",
      jazz = "#4575b4",
      metal = "#7f3b08"
    ),
    name = "Genre"
  ) +
  coord_fixed(1.3, xlim = c(-180, 180), ylim = c(-60, 85)) +
  labs(
    title = "Artist Locations by Inferred Genre",
    subtitle = "Geographic pattern of artists assigned to rock, pop, blues, jazz, or metal based on artist terms.",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#e6f0f6"),
    panel.grid = element_line(color = "gray90"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )

# Save the plot to a PNG file
print(gg)
```

```{r}
# Load required packages
library(ggplot2)
library(maps)

# Read the music dataset
music <- read.csv("data/music.csv", stringsAsFactors = FALSE)

# Convert song.year and coordinate columns to numeric
music$song.year <- as.numeric(as.character(music$song.year))
music$artist.latitude <- as.numeric(as.character(music$artist.latitude))
music$artist.longitude <- as.numeric(as.character(music$artist.longitude))

# Filter to artists with usable coordinates and nonzero song.year
year_artists <- subset(music,
                       song.year != 0 & !is.na(song.year) &
                       !(artist.latitude == 0 & artist.longitude == 0) &
                       !is.na(artist.latitude) & !is.na(artist.longitude))

# Load world map data
world_map <- map_data("world")

# Build the plot
gg <- ggplot() +
  geom_polygon(
    data = world_map,
    aes(x = long, y = lat, group = group),
    fill = "gray95",
    color = "gray70",
    linewidth = 0.2
  ) +
  geom_point(
    data = year_artists,
    aes(x = artist.longitude, y = artist.latitude, color = song.year),
    alpha = 0.75,
    size = 1.8
  ) +
  scale_color_viridis_c(option = "plasma", name = "Song Year") +
  coord_fixed(1.3, xlim = c(-180, 180), ylim = c(-60, 85)) +
  labs(
    title = "Artist Locations Colored by Song Year",
    subtitle = "Shows artists with usable coordinates and nonzero song year; geographic coverage is limited to available geocoded artists.",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#e6f0f6"),
    panel.grid = element_line(color = "gray90"),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )

# Save the plot to a PNG file
print(gg)
```
